\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-07-21 tex/methods.tex
update sections on structural stability

The quality of interest in this paper is robustness, which is related to the concept of structural stability (see \ref{sec:dynrobustandstructstab}), whose evaluation requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters, its rate functions, or environmental constraints that restrict it to a subset of its basins of attraction. We mean to refer to perturbations to the structure of the system itself as determined by the strengths of the couplings between the components and not only to perturbations of the state vector at a given point in time. It is justified to consider resampling elements of $A$ to generate $A'$ as a proxy for resampling elements of $\vec{p}$ to produce $\vec{p}\,'$ if any matrix $A$ can be obtained for some $F_i$, $\vec{p}$ and $\vec{x}^0$. This holds for the $F_i$ defining the Lotka-Volterra model. This is due to the fact that for a specification of non-zero real numbers for the components of $\vec{x}^0$ and any real numbers for the components of $a_{ij}$, there is a choice of parameters $\vec{p}$ given by $b_{ij} = \frac{a_{ij}}{x_i^0}$ and $r_i = - \sum_{j=1}^n \frac{x_j^0}{x_i^0} a_{ij}$ that generates those particular $a_{ij}$ as the Jacobian matrix of the dynamical system. Checking this property of the domain of realizability of the Jacobian can be done for ensembles of systems other than the Lotka-Volterra ensemble. For arbitrary biochemical reaction and gene regulatory networks, this property is likely to hold so long as not too many types of transformations are constrained from possibility. For example, a simplified version of the general form of the gene regulatory network model presented in \ref{fig:biomodelexamples} center panel is given by the system
\begin{equation}
\frac{dg_i}{dt} = \sum_{j=1}^N k_{ij} g_j,
\end{equation}
with one parameter $k_{ij} \in \mathbb{R}$ for every pair $(g_i,g_j)$ of genes. The Jacobian of this system is $A_{ij} = k_{ij}$, and, therefore, sampling parameters of the model is precisely equivalent to sampling elements of the Jacobian.
% For reaction networks, we demonstrate an ensemble from which arbitrary Jacobians can arise in \ref{sec:reactionnetjacobian}.

@@

Structural stability is usually formulated in terms of the space of all smooth vector fields endowed with the $C^1$ topology.  We did not want to do this for two reasons.  Theoretically, it is not clear how one would impose a probability measure on the set of all vector fields.  Practically, one does not expect that an environmental fluctuation will often change a dynamical system into any other arbitrary system but that it may introduce a change into a \emph{nearby} system that shares many features with the original.  Hence we formulate our definition in terms of a finite-dimensional, if still large, parameter space.

Structural stability is defined in terms of topological conjugacy which is a global condition ensuring that all features of the phase space are preserved.  We, however, are only interested in the portion of the state space near a certain fixed point and are not concerned with perturbations that might change the character of the phase space at points far away from the region of interest as long as they do not affect the neighborhood of the fixed point.  Thus, rather than being formulated globally, our definition is given locally in terms of the stability of a given fixed point.

>>>>

% The quality of interest in this paper is robustness, which is related to the concept of structural stability (see \ref{sec:dynrobustandstructstab}), whose
The evaluation of \ref{eq:robustnessonjacobians} requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters, its rate functions, or environmental constraints that restrict it to a subset of its basins of attraction. We mean to refer to perturbations to the structure of the system itself as determined by the strengths of the couplings between the components and not only to perturbations of the state vector at a given point in time. It is justified to consider resampling elements of $A$ to generate $A'$ as a proxy for resampling elements of $\vec{p}$ to produce $\vec{p}\,'$ if any matrix $A$ can be obtained for some $F_i$, $\vec{p}$ and $\vec{x}^0$. This holds for the $F_i$ defining the Lotka-Volterra model. This is due to the fact that for a specification of non-zero real numbers for the components of $\vec{x}^0$ and any real numbers for the components of $a_{ij}$, there is a choice of parameters $\vec{p}$ given by $b_{ij} = \frac{a_{ij}}{x_i^0}$ and $r_i = - \sum_{j=1}^n \frac{x_j^0}{x_i^0} a_{ij}$ that generates those particular $a_{ij}$ as the Jacobian matrix of the dynamical system. Checking this property of the domain of realizability of the Jacobian can be done for ensembles of systems other than the Lotka-Volterra ensemble. For arbitrary biochemical reaction and gene regulatory networks, this property is likely to hold so long as not too many types of transformations are constrained from possibility. For example, a simplified version of the general form of the gene regulatory network model presented in \ref{fig:biomodelexamples} center panel is given by the system
\begin{equation}
\frac{dg_i}{dt} = \sum_{j=1}^N k_{ij} g_j,
\end{equation}
with one parameter $k_{ij} \in \mathbb{R}$ for every pair $(g_i,g_j)$ of genes. The Jacobian of this system is $A_{ij} = k_{ij}$, and, therefore, sampling parameters of the model is precisely equivalent to sampling elements of the Jacobian.
% For reaction networks, we demonstrate an ensemble from which arbitrary Jacobians can arise in \ref{sec:reactionnetjacobian}.

@@

Structural stability is usually formulated in terms of the space of all smooth vector fields endowed with the $C^1$ topology.  We did not want to do this for two reasons.  Theoretically, it is not clear how one would impose a probability measure on the set of all vector fields.  Practically, we do not consider environmental fluctuations that introduce a smooth simultaneous change to all possible interactions among system variables, but rather those that introduce a change in one or changes in a few interactions among system variables.  Hence we formulate our definition in terms of a finite-dimensional, if still large, space of possible systems.

Structural stability is defined in terms of topological conjugacy which is a global condition ensuring that all features of the phase space are preserved.  We, however, are only interested in the portion of the state space near a particular fixed point and are not concerned with perturbations that might change the character of the phase space at points far away from the region of interest as long as they do not affect the neighborhood of the fixed point.  Thus, rather than being formulated globally, our definition is given locally in terms of the stability of a given fixed point.

### 2015-07-21 tex/overview.tex
motivate the inclusion of symmetries
update sections on structural stability

In the general case of a dynamical system with $n$ components, where the components may be concentrations of chemical species, genes, or biological species, we have an $n$-dimensional vector of state variables or observables $(x_1(t), \ldots x_n(t)) = \vec{x}(t)
$
whose components are solutions to the arbitrary first order system
\begin{equation}\label{eq:eom}
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t), \vec{p}), \; (i=1,\ldots,n)
\end{equation}
where $F=\{F_1,\ldots,F_i,\ldots,F_n\}$ represent, potentially nonlinear, functions of the given vector of state variables and $\vec{p} \in \mathbb{R}^{n'}$ is the vector of $n'$ parameters of the $F_i$. These parameters typically represent reaction rates or interaction strengths in chemical, gene-regulatory and ecological networks. For example, in the Lotka-Volterra model in \ref{fig:biomodelexamples}, $\vec{x} = (x_1, \ldots, x_n)$, $\vec{p}=(r_1,\ldots,r_n,b_{11},\ldots,b_{nn})$, and $F_i = r_i x_i + \sum_{j=1}^{n} b_{ij} x_i x_j$. The set of all dynamical systems, $\mathcal{D}$, for a given number of state variables $n$ and a given number of parameters $n'$, is then
\begin{equation}\label{eq:setdynsys}
\mathcal{D'} = \{ (F(-,\vec{p}),\vec{p}) | F \colon \mathbb{R}^n \times \mathbb{R}^{n'} \rightarrow \mathbb{R}^n, \vec{p} \in \mathbb{R}^s \}.
\end{equation}
The space of all $F$ is too large to precisely define probabilities over, so we consider various subspaces $\mathcal{D} \subset \mathcal{D}'$. In some cases these may be given by parametrized models with a single fixed form of $F$ and then only $\vec{p}$ is allowed to vary.

@@

\section{Evolutionary processes sampling dynamical systems}
In the course of biological evolution, the parameter values $\vec{p}$, form of the functions $F$, and environmental conditions restricting access to the basins associated to different fixed points $\vec{x}^0$ corresponding to all different types of networks considered in \ref{fig:biomodelexamples} are subject to, potentially drastic, modifications due to environmental fluctuations. The stochastic process by which these modifications occur induces one on the set of fixed points that results in the assignment of a probability $P(f^T)$ to each history of length $T$, $f^T = ( f_1,f_2,\ldots,f_T ) \in \mathcal{F}^T$ \cite{RobertM.Gray130}.
% $$
% P(f',t+1 | f,t) = \mathcal{T}_{\mathcal{F}}(f',f)
% $$
% for all $f,f' \in \mathcal{F}$.
These dynamics induce a stochastic process on the equivalence classes of fixed points $\mathcal{F}/{\sim}$ indexed by Jacobian matrices given by
$$
P([A]^T) = \sum_{ \{ f^T \mid f_i^T \in [A]_i^T \} } P(f^T)
$$
for each history of length $T$, $[A]^T = ( [A_1], [A_2], \ldots, [A_T] ) \in (\mathcal{F}/{\sim})^T$
% $$
% P([A'],t+1 | [A],t) = \mathcal{T}_{\mathcal{F}/{\sim}}([A'],[A])
% $$
% for all $A,A' \in \mathcal{F}/{\sim}$
as visualized in \ref{fig:robustnessprocess}B. These changes can alter the stability of a given system, thereby inducing an even more coarse-grained stochastic process on the stable regions of the space of fixed points given by
$$
P(s^T) = \sum_{ \{ A^T \mid \mathcal{S}(A_i^T) = s_i^T \} } P(A^T)
$$
for each history of length $T$, $s^T = ( s_1, s_2, \ldots, s_T ) \in \{0,1\}^T$ where $s=1 \equiv S,\, s=0 \equiv U$ as visualized in \ref{fig:robustnessprocess}A and B. In order to model this we consider a process whereby perturbations applied to a given dynamical system and fixed point associated to a stable Jacobian matrix $A$ lead to another Jacobian matrix $A'$.  This corresponds to an ensemble of fixed points of dynamical systems where each model in the ensemble may otherwise be defined in terms of a different collection of rate functions $F'$, vector of parameters ${\vec{p}}\,'$, or environmental conditions restricting access to $\vec{x}^{0'}$ (\refsupp{}). We then ask what is the probability, given $A$ is a stable matrix, that $A'$ is also a stable matrix. This quantifies the intuitive statement that, over evolutionary timescales, it is not enough for a system to be stable, but rather that it have a reasonably high probability of remaining stable over continguous timeframes. In terms of a history $s^T$ over the states specifying the stability property, $\mathcal{S}$, this is given by
$$
r = \frac{\sum_{t=0}^{T-1} s_{t+1}^T s_{t}^T}{\sum_{t=0}^{T-1} s_{t}^T}.
$$
If the process $P(s^T)$ has limits $\mu(s) = \lim_{t,T \rightarrow \infty} P(s_t^T=s)$ and $\tau(s' \mid s) = \lim_{t,T \rightarrow \infty} P(s_{t+1}^T = s' \mid s_t^T = s)$ for $s,s' \in \{0,1\}$
%stationary probability $P^{\infty}(s)$ for $s \in \mathcal{S}$
, the expectation of $r$ as $T \rightarrow \infty$ is approximated by
\begin{equation}\label{eq:robustnessfromstoch}
R(S,\mu,\tau) = E[r] \approx \frac{\sum_{s,s'} \mu(s) \tau(s' \mid s)}{\sum_{s=1} \mu(s)}.
\end{equation}
This conditional probability corresponds to the parameter $R$ of the two-state process depicted in \ref{fig:robustnessprocess}A, and it is what we refer to as dynamical robustness.

In terms of Jacobian matrices $A$ and $A'$, if the underlying process $P([A]^T)$ has analogous limits
$$\mu(A) = \lim_{t,T \rightarrow \infty} P(A_t^T=A)$$
and
$$\tau(A' \mid A) = \lim_{t,T \rightarrow \infty} P(A_{t+1}^T = A' \mid A_t^T = A)$$
for $A,A' \in \mathbb{R}^{n \times n}$, \ref{eq:robustnessfromstoch} becomes
\begin{equation}
R(S,\mu,\tau) = \frac{\int_{A,A' \in \mathbb{R}^{n \times n}} d\mu(A) d\tau(A' \mid A) \mathcal{S}(A) \mathcal{S}(A')}{\int_{A \in \mathbb{R}^{n \times n}} d\mu(A) \mathcal{S}(A)}.
\end{equation}

@@

Symmetries with respect to some property of the system are characterized by the ability to interchange these modules or their connectivity without changing that property. Two of these three intrinsic symmetries of $\hier$ are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these latter symmetries applied to a specific interaction graph.

>>>>

In the general case of a dynamical system with $n$ components, where the components may be concentrations of chemical species, genes, or biological species, we have an $n$-dimensional vector of state variables or observables $(x_1(t), \ldots x_n(t)) = \vec{x}(t)
$
whose components are solutions to the arbitrary first order system
\begin{equation}\label{eq:eom}
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t), \vec{p}), \; (i=1,\ldots,n)
\end{equation}
where $F=\{F_1,\ldots,F_i,\ldots,F_n\}$ represent, potentially nonlinear, functions of the given vector of state variables and $\vec{p} \in \mathbb{R}^{n'}$ is the vector of $n'$ parameters of the $F_i$. These parameters typically represent reaction rates or interaction strengths in chemical, gene-regulatory and ecological networks. For example, in the Lotka-Volterra model in \ref{fig:biomodelexamples}, $\vec{x} = (x_1, \ldots, x_n)$, $\vec{p}=(r_1,\ldots,r_n,b_{11},\ldots,b_{nn})$, and $F_i = r_i x_i + \sum_{j=1}^{n} b_{ij} x_i x_j$. The set of all dynamical systems, $\mathcal{D}$, for a given number of state variables $n$ and a given number of parameters $n'$, is then
\begin{equation}\label{eq:setdynsys}
\mathcal{D'} = \{ (F(-,\vec{p}),\vec{p}) | F \colon \mathbb{R}^n \times \mathbb{R}^{n'} \rightarrow \mathbb{R}^n, \vec{p} \in \mathbb{R}^s \}.
\end{equation}
The space of all possible $F$ may be too large to precisely define probabilities over, in which case it is possible to consider various subspaces $\mathcal{D} \subset \mathcal{D}'$. In some cases these may even be given by parametrized models with a single fixed form of $F$ and then varying $\vec{p}$ alone would be sufficient to vary within that class of dynamical systems.

@@

\section{Evolutionary processes sampling dynamical systems}
In the course of biological evolution, the parameter values $\vec{p}$, form of the functions $F$, and environmental conditions restricting access to the basins associated to different fixed points $\vec{x}^0$ corresponding to all different types of networks considered in \ref{fig:biomodelexamples} are subject to, potentially drastic, modifications due to environmental fluctuations. The stochastic process by which these modifications occur induces one on the set of fixed points that results in the assignment of a probability $P(f^T)$ to each history of length $T$, $f^T = ( f_1,f_2,\ldots,f_T ) \in \mathcal{F}^T$ \cite{RobertM.Gray130}.
% $$
% P(f',t+1 | f,t) = \mathcal{T}_{\mathcal{F}}(f',f)
% $$
% for all $f,f' \in \mathcal{F}$.
These dynamics induce a stochastic process on the equivalence classes of fixed points $\mathcal{F}/{\sim}$ indexed by Jacobian matrices given by
$$
P([A]^T) = \sum_{ \{ f^T \mid f_i^T \in [A]_i^T \} } P(f^T)
$$
for each history of length $T$, $[A]^T = ( [A_1], [A_2], \ldots, [A_T] ) \in (\mathcal{F}/{\sim})^T$
% $$
% P([A'],t+1 | [A],t) = \mathcal{T}_{\mathcal{F}/{\sim}}([A'],[A])
% $$
% for all $A,A' \in \mathcal{F}/{\sim}$
as visualized in \ref{fig:robustnessprocess}B. These changes can alter the stability of a given system, thereby inducing an even more coarse-grained stochastic process on the stable regions of the space of fixed points given by
$$
P(s^T) = \sum_{ \{ A^T \mid \mathcal{S}(A_i^T) = s_i^T \} } P(A^T)
$$
for each history of length $T$, $s^T = ( s_1, s_2, \ldots, s_T ) \in \{0,1\}^T$ where $s=1 \equiv S,\, s=0 \equiv U$ as visualized in \ref{fig:robustnessprocess}A and B. In order to model this we consider a process whereby perturbations applied to a given dynamical system and fixed point associated to a stable Jacobian matrix $A$ lead to another Jacobian matrix $A'$.  This corresponds to an ensemble of fixed points of dynamical systems where each model in the ensemble may otherwise be defined in terms of a different collection of rate functions $F'$, vector of parameters ${\vec{p}}\,'$, or environmental conditions restricting access to $\vec{x}^{0'}$ (\refsupp{}). We then ask what is the probability, given $A$ is a stable matrix, that $A'$ is also a stable matrix. This quantifies the intuitive statement that, over evolutionary timescales, it is not enough for a system to be stable, but rather that it have a reasonably high probability of remaining stable over contiguous timeframes. In terms of a history $s^T$ over the states specifying the stability property, $\mathcal{S}$, this is given by
$$
r = \frac{\sum_{t=0}^{T-1} s_{t+1}^T s_{t}^T}{\sum_{t=0}^{T-1} s_{t}^T}.
$$
If the process $P(s^T)$ has limits $\mu(s) = \lim_{t,T \rightarrow \infty} P(s_t^T=s)$ and $\tau(s' \mid s) = \lim_{t,T \rightarrow \infty} P(s_{t+1}^T = s' \mid s_t^T = s)$ for $s,s' \in \{0,1\}$
%stationary probability $P^{\infty}(s)$ for $s \in \mathcal{S}$
, the expectation of $r$ as $T \rightarrow \infty$ is approximated by
\begin{equation}\label{eq:robustnessfromstoch}
R(S,\mu,\tau) = E[r] \approx \frac{\sum_{s,s'} \mu(s) \tau(s' \mid s)}{\sum_{s=1} \mu(s)}.
\end{equation}
This conditional probability corresponds to the parameter $R$ of the two-state process depicted in \ref{fig:robustnessprocess}A, and it is what we refer to as dynamical robustness.

Conceptually, one might consider the quantity defined by \ref{eq:robustnessfromstoch} to be related to structural stability \cite{Smale1967}, but its definition differs in several important ways (see \ref{sec:dynrobustandstructstab}). Briefly, structural stability is a qualitative, global property that distinguishes dynamical systems whose associated vector field topologies are invariant to \emph{any} sufficiently small perturbation from those whose topologies undergo some change under any such perturbations. Dynamical robustness, on the other hand, is a quantitative, local property that specifies the probability of the stability of a fixed point undergoing a transformation from stable to unstable when the system is subjected to, potentially large, perturbations in one or a few interactions among system variables.

In terms of Jacobian matrices $A$ and $A'$, if the underlying process $P([A]^T)$ has analogous limits
$$\mu(A) = \lim_{t,T \rightarrow \infty} P(A_t^T=A)$$
and
$$\tau(A' \mid A) = \lim_{t,T \rightarrow \infty} P(A_{t+1}^T = A' \mid A_t^T = A)$$
for $A,A' \in \mathbb{R}^{n \times n}$, \ref{eq:robustnessfromstoch} becomes
\begin{equation}\label{eq:robustnessonjacobians}
R(S,\mu,\tau) = \frac{\int_{A,A' \in \mathbb{R}^{n \times n}} d\mu(A) d\tau(A' \mid A) \mathcal{S}(A) \mathcal{S}(A')}{\int_{A \in \mathbb{R}^{n \times n}} d\mu(A) \mathcal{S}(A)}.
\end{equation}

@@

Symmetries with respect to some property of the system are characterized by the ability to interchange these modules or their connectivity without changing that property. Two of these three intrinsic symmetries of $\hier$ are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these latter symmetries applied to a specific interaction graph. The collections of networks that have equivalent dynamical robustness characterized by its symmetries allow for a classification of neutral networks with respect to any process in which maximal robustness is selected for.
